A Comparison of Parallel Graph Processing Benchmarks

### 2017-02-16 papers/publication/paper-pollard.tex
minor reorg and todo
small move
Updated paper with more exposition and todos.

\documentclass{llncs}
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{hyperref}
\usepackage{cite} % Multiple citations in one \cite command
\usepackage{pgfplotstable}
\usepackage{array} % Required for pgfplotstable dec sep align
\pgfplotsset{compat=1.12}
\usepackage{booktabs}

@@

@@

\section{Experimental Setup}

@@

We consider three algorithms though not all algorithms are available on all systems.

\begin{enumerate}
	\item Breadth First Search (BFS)
	\item Single Source Shortest Paths (SSSP)
	\item PageRank
\end{enumerate}

The canonical performance leaderboard for parallel graph processing is the Graph500 \cite{Murphy:2010:Graph500}. The advantage of the Graph500 is it provides standardized measurement specifications and dataset generation. The primary drawback with the Graph500 is it measures a single algorithm: BFS.

Our work aims to add similar rigor to other graph algorithms by borrowing heavily from the Graph500 specification. The Graph500 Benchmark 1 (``Search'') is concerned with two kernels: the creation of a graph data structure from an edge list stored in RAM and the actual BFS\footnote{For a complete specification, see \url{http://graph500.org/specifications}}. We run the BFS using 32 random roots.

One straightforward extension to BFS and our second algorithm is the Single-Source Shortest Paths algorithm (SSSP). We use the same graph and the same source vertices as in BFS.

The third algorithm is PageRank. These three algorithms are used because of their popularity; most libraries provide reference implementations.

@@

\subsection{Datasets}
We use the Graph500 synthetic graph generator which creates a Kronecker graph \cite{Leskovec:2010:Kronecker} with initial parameters of $A = 0.57, B = 0.19, C = 0.19,$ and $D = 1-(A+B+C) = 0.05$.

The Graphalytics results in Table~\ref{tab:graphalytics} were performed on the Dota-League dataset. This dataset contains 61,670 vertices and 50,870,313 edges. This dataset is sourced from the Game Trace Archive\cite{Guo:2012:GTA} and modified for Graphalytics\footnote{This dataset is available at \url{https://atlarge.ewi.tudelft.nl/graphalytics/}.}.

\section{Performance Analysis}

@@

Table~\ref{tab:epg-perf}, in contrast, gives an overview of the performance results of running graphalytics and our approach using the same dataset [TODO: Get the Dota-League Dataset working with eas-parallel-graph and replace Table~\ref{epg-perf}]. Table~\ref{tab:epg-perf} shows the average time across 32 roots, whereas Graphalytics by default shows results for a single root.

Further detail is shown in Figs.~\ref{fig:bfs-time} and \ref{fig:sssp-time}, giving box plots for the runtime distributions. However, even these do not give a complete picture of performance. Both of these show GraphMat to not be highly performant. However, a likely explanation is that GraphMat's underlying computation model (sparse matrix operations) paired the increased overhead of GraphMat's doubly-compressed sparse row (DSCR) graph representation is more conducive to larger-scale graphs.

@@

\begin{figure}
	\centering
	\caption{The $y$-axis is logarithmic. GraphBIG reads in the file and generates the data structure simultaneously so is omitted.}
	\begin{minipage}{0.48\linewidth}
		\includegraphics[width=\linewidth, trim=0 36pt 18pt 0, clip]{graphics/bfs_time.pdf}
	\end{minipage}
	\begin{minipage}{0.48\linewidth}
		\includegraphics[width=\linewidth, trim=0 36pt 18pt 0, clip]{graphics/bfs_dsc.pdf}
	\end{minipage}
	\label{fig:bfs-time}
\end{figure}

\begin{figure}
	\centering
	\begin{minipage}{0.59\linewidth}
		\includegraphics[width=\linewidth, trim=0 36pt 18pt 0, clip]{graphics/sssp_time.pdf}
	\end{minipage}
	\begin{minipage}{0.365\linewidth}
		\includegraphics[width=\linewidth, trim=0 36pt 18pt 0, clip]{graphics/sssp_dsc.pdf}
	\end{minipage}
	\caption{The $y$-axis is logarithmic. Both PowerGraph and GraphBIG construct their data structures at the same time as they read in the file.}
	\label{fig:sssp-time}
\end{figure}

The behavior of PageRank is slightly different. As with SSSP and BFS, the GAP Benchmark Suite is the fastest, and it also requires the fewest iterations. While care was taken to ensure the same stopping criterion is met, the particular implementation methods make it impossible to ensure the same computations are taking place. In general, we measured $\sum_{k=1}^{n} |p_k^{(i)} - p_k^{(i-1)}|$ where $i$ is the iteration and $n$ is the number of vertices, both the vertex-centric approaches, PowerGraph and GraphMat, take more iterations.

\begin{figure}
	\centering
	\begin{minipage}{0.48\linewidth}
		\includegraphics[width=\linewidth, trim=0 18pt 18pt 0, clip]{graphics/pr_time.pdf}
	\end{minipage}
	\begin{minipage}{0.48\linewidth}
		\includegraphics[width=\linewidth, trim=0 18pt 18pt 0, clip]{graphics/pr_iters.pdf}
	\end{minipage}
	\caption{The $y$-axis is logarithmic only for the left figure. GraphMat continues to run until none of the vertices' ranks change. GraphMat iterations are measured differently because of the ``think like a vertex'' paradigm and runs until none of the vertices' ranks change. For the others, we use the stopping condition that the sum of the changes in the weights is no more than $6 \times 10^{-8}$, or approximately machine epsilon for single precision floating point numbers.}
	\label{fig:pr}
\end{figure}

@@

The challenge in comparing iteration counts for PageRank underscores an important challenge for any comparison of graph processing systems. The assumptions under which the various platforms operate can have a dramatic effect on the program. For example, the GAP Benchmark Suite stores vertex weights as 32-bit integers. However, other systems store them as floating point numbers. This may affect performance in addition to runtime behavior in cases where weights like $0.2$ are cast to $0$. Similarly, how a grpah is represented in the system (e.g. weighed or directed) may have performance and algorithmic implications but is not always readily apparent.

In Fig.~\ref{fig:bfs-scaling}, we see BFS scalability\dots
\begin{figure}
	\centering
	\begin{minipage}{0.55\linewidth}
		\includegraphics[width=\linewidth, trim=0 18pt 18pt 0, clip]{graphics/bfs_speedup.pdf}
	\end{minipage}
	\begin{minipage}{0.425\linewidth}
		\includegraphics[width=\linewidth, trim=0 18pt 18pt 0, clip]{graphics/bfs_ss.pdf}
	\end{minipage}
	\caption{Both figures use the kronecker graph with $2^{20}$ vertices with an average of 16 edges per vertex.}
	\label{fig:bfs-scaling}
\end{figure}

>>>>

\documentclass{llncs}
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{hyperref}
\usepackage{cite} % Multiple citations in one \cite command
\usepackage{pgfplotstable}
\usepackage{array} % Required for pgfplotstable dec sep align
\pgfplotsset{compat=1.12}
\usepackage{booktabs}
\usepackage{todonotes}

@@

Our approach differs from the previously mentioned appproaches because it focuses on the depth of analysis rather than the breadth via diversity of datasets or graph processing systems. We analyze  power and energy consumption in addition to details such as the number of iterations or the time to construct the data structures. Additionally, our approach requires little knowledge of the inner workings of each system; the data are collected using either sampling of model specific registers (for power) or parsing log files (for timing).

@@

\section{Experimental Setup}
We refer to a graphs's \emph{scale} when describing the size of the graph. Specifically, a graph with scale $S$ has $2^S$ vertices. The datasets are described further in Sect.~\ref{sec:datasets}. We measure strong scaling and speedup by varying the number of threads from one to the total number of threads on the reasearch machine, 72.

Each experiment uses 32 roots per graph. As with the Graph500, each root is selected to have a degree greater than 1. In the case of PageRank, we simply run PageRank 32 times.

When possible, we measure the data set construction time as the time to translate from the unstructured file data in RAM to the graph representation on which the algorithm can be performed. This is not possible for PowerGraph and GraphBIG because they read in the input file and build a graph simultaneously. The Graph500 builds the graph once and performs the BFS on each root in turn, speeding up experimentation but only giving a single data point for the data set construction.

We plot many of the results as box plots with an implied 32 datapoints per box.

@@

We consider three algorithms: Breadth First Search (BFS), Single Source Shortest Paths (SSSP), and PageRank, though not all algorithms are implemented on all systems.

We select BFS because the canonical performance leaderboard for parallel graph processing is the Graph500 \cite{Murphy:2010:Graph500}. The advantage of the Graph500 is it provides standardized measurement specifications and dataset generation. The primary drawback with the Graph500 is it measures a single algorithm.

Our work aims to add similar rigor to other graph algorithms by borrowing heavily from the Graph500 specification. The Graph500 Benchmark 1 (``Search'') is concerned with two kernels: the creation of a graph data structure from an edge list stored in RAM and the actual BFS\footnote{For a complete specification, see \url{http://graph500.org/specifications}}. We run the BFS using 32 random roots with the exception of PowerGraph which doesn't provide an reference implementation of BFS in its toolkits.

We select SSSP because of the straightforward extension from BFS; we need not modify the graph nor the root vertices from BFS.

PageRank is selected because of its popularity; most libraries provide reference implementations. One challenge with using PageRank is the stopping criterion; this is addressed in Sect.~\ref{sec:perf}. Verification of the PageRanks is beyond the scope of this paper though this may explain some of the large performance discrepancies.

This approach is not specific to a particular algorithm; measuring the execution time, data structure construction time, and power consumption can be easily extended to other algorithms.

@@

\subsection{Datasets}\label{sec:datasets}
We use the Graph500 synthetic graph generator which creates a Kronecker graph \cite{Leskovec:2010:Kronecker} with initial parameters of $A = 0.57, B = 0.19, C = 0.19,$ and $D = 1-(A+B+C) = 0.05$ and set the average degree of a vertex as 16. Hence, a Kronecker graph with scale $S$ has $2^S$ vertices and approximately $16 \times 2^S$ edges.

The Graphalytics results in Table~\ref{tab:graphalytics} were performed on the Dota-League dataset. This dataset contains 61,670 vertices and 50,870,313 edges. This dataset comes from the Game Trace Archive\cite{Guo:2012:GTA} and is modified for Graphalytics\footnote{This dataset is available at \url{https://atlarge.ewi.tudelft.nl/graphalytics/}.}. This dataset is useful because it is both weighted and more dense than the usual real-world dataset with an average out-degree of 824.

\section{Performance Analysis}\label{sec:perf}

@@

\todo{rework this. It would be nice to compare Graphalytics and our approach on the real-world dataset.}
Table~\ref{tab:epg-perf}, in contrast, gives an overview of the performance results of running graphalytics and our approach using the same dataset [TODO: Get the Dota-League Dataset working with easy-parallel-graph and replace Table~\ref{epg-perf}]. Table~\ref{tab:epg-perf} shows the average time across 32 roots, whereas Graphalytics by default shows results for a single root.

Figures~\ref{fig:bfs-time} and \ref{fig:sssp-time} show more detail, with box plots for the runtime distributions. However, even these do not give a complete picture of performance. Both of these show GraphMat to not be highly performant. One explanation is GraphMat's underlying computation model (sparse matrix operations) paired the increased overhead of GraphMat's doubly-compressed sparse row (DSCR) graph representation is more conducive to larger-scale graphs. There is less discrepancy between the runtimes of SSSP (between 0.1 and 1.7 seconds) compared to BFS (0.01 and 1.7 seconds) but GAP is the clear winner in both cases. The data structure construction times for GAP and GraphMat are consistent; in both cases the platforms create the same data structure for both algorithms.

These results are consistent with \cite{Sundaram:2015:GraphMat} which lists GraphMat as more performant than PowerGraph in SSSP.

@@

\todo{Make the text bigger for these}
\begin{figure}
	\centering
	\begin{minipage}{0.48\linewidth}
		\includegraphics[width=\linewidth, trim=0 36pt 18pt 0, clip]{graphics/bfs_time.pdf}
	\end{minipage}
	\begin{minipage}{0.48\linewidth}
		\includegraphics[width=\linewidth, trim=0 36pt 18pt 0, clip]{graphics/bfs_dsc.pdf}
	\end{minipage}
	\caption{The $y$-axes are logarithmic. The left box plot shows the time to compute BFS on 32 random roots while the right plot shows the times to construct the graph for each system. The Graph500 only constructs its graph once. GraphBIG reads in the file and generates the data structure simultaneously so is omitted.}
	\label{fig:bfs-time}
\end{figure}

\begin{figure}
	\centering
	\begin{minipage}{0.59\linewidth}
		\includegraphics[width=\linewidth, trim=0 36pt 18pt 0, clip]{graphics/sssp_time.pdf}
	\end{minipage}
	\begin{minipage}{0.365\linewidth}
		\includegraphics[width=\linewidth, trim=0 36pt 18pt 0, clip]{graphics/sssp_dsc.pdf}
	\end{minipage}
	\caption{The $y$-axes is logarithmic. The left box plot shows the time to compute the SSSP starting at the same 32 roots as Fig.~\ref{fig:bfs-time}. Both PowerGraph and GraphBIG construct their data structures at the same time as they read in the file.}
	\label{fig:sssp-time}
\end{figure}

The behavior of PageRank is slightly different. As with SSSP and BFS, the GAP Benchmark Suite is the fastest but it also requires the fewest iterations.

We attempt to use similar stopping criteria for each system, but GraphMat executes until no vertices change rank; effectively its stopping criterion requires the $\infty$-norm be less than machine epsilon. This could account for the increased number of iterations. We adjusted the other systems to use $\sum_{k=1}^{n} |p_k^{(i)} - p_k^{(i-1)}| < \epsilon $ as the stopping criteria, where $i$ is the iteration and $n$ is the number of vertices. We use $\epsilon = 6 \times 10^{-8}$ because this value is approximately machine epsilon for a single precision floating point number. This is done with the goal of making the stoppping criteria as similar as possible.

Furthermore, the logarithmic scale in Figs.~\ref{fig:pr} and~\ref{fig:bfs-time} belies the actual variance in results. Each platform in Fig.~\ref{fig:pr} has a standard deviation comparable to that of

%> sd(pr_time$Time[pr_time$Sys == "PowerGraph"])
%[1] 0.4654061
%> sd(sssp_time$Time[sssp_time$Sys == "PowerGraph"])
%[1] 0.1475608
%> sd(pr_time$Time[pr_time$Sys == "GAP"])
%[1] 0.01957317
%> sd(sssp_time$Time[sssp_time$Sys == "GAP"])
%[1] 0.01608644
%> sd(pr_time$Time[pr_time$Sys == "GraphMat"])
%[1] 0.9305763
%> sd(sssp_time$Time[sssp_time$Sys == "GraphMat"])
%[1] 0.02191714
%> sd(pr_time$Time[pr_time$Sys == "GraphBIG"])
%[1] 3.308351
%> sd(sssp_time$Time[sssp_time$Sys == "GraphBIG"])
%[1] 0.2165655

\begin{figure}
	\centering
	\begin{minipage}{0.48\linewidth}
		\includegraphics[width=\linewidth, trim=0 18pt 18pt 0, clip]{graphics/pr_time.pdf}
	\end{minipage}
	\begin{minipage}{0.48\linewidth}
		\includegraphics[width=\linewidth, trim=0 18pt 18pt 0, clip]{graphics/pr_iters.pdf}
	\end{minipage}
	\caption{The $y$-axis is logarithmic only for the left figure. GraphMat continues to run until none of the vertices' ranks change. For the others, we use the stopping condition that the sum of the changes in the weights is no more than $6 \times 10^{-8}$, or approximately machine epsilon for single precision floating point numbers.}
	\label{fig:pr}
\end{figure}
\todo{Check this again---that's a huge discrepancy!}

@@

The challenge in comparing iteration counts for PageRank underscores an important challenge for any comparison of graph processing systems. The assumptions under which the various platforms operate can have a dramatic effect on the program. For example, the GAP Benchmark Suite stores vertex weights as 32-bit integers. However, other systems store them as floating point numbers. This may affect performance in addition to runtime behavior in cases where weights like $0.2$ are cast to $0$. Similarly, how a graph is represented in the system (e.g. weighed or directed) may have performance and algorithmic implications but is not always readily apparent.

In Fig.~\ref{fig:bfs-scaling}, we see BFS scalability\dots

\begin{figure}
	\centering
	\begin{minipage}{0.9\linewidth}
		\includegraphics[width=\linewidth, trim=0 18pt 18pt 0, clip]{graphics/bfs_speedup.pdf}
	\end{minipage}
	\begin{minipage}{0.9\linewidth}
		\includegraphics[width=\linewidth, trim=0 18pt 18pt 0, clip]{graphics/bfs_ss.pdf}
	\end{minipage}
	\caption{The top figure has both axes logarithmic. Linear speedup is. GraphMat dips below 1 because it is slower for $2$ threads than for 1.}
	\label{fig:bfs-scaling}
\end{figure}
\todo{Rerun GraphMat on scale 22 with 2 threads. Surprising that it slows down}
